Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-08-26 ms.Rnw
Added/removed paragraph, and other minor edits

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{\hspace{6mm} ---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{D}
\newcommand\code[1]{\texttt{#1}}

@@

\begin{abstract}
Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions for stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
% \eic{EDIT IN THE DIRECTION OF THE PREVIOUS SENTENCE.}
As a case study, we consider a cholera epidemic in Haiti.
The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission, eliminated in 2019 with the help of vaccination and other public health measures.
We study three models developed by expert teams to advise on vaccination policies.
We assess methods used for fitting and evaluating these models, leading to recommendations for future studies.
% Methods that lead to better statistical fit include benchmarking mechanistic models against associative statistical models and using state-of-the art tools designed to fit mechanistic models.
Better statistical fit facilitates interpretation of the fitted model, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
\end{abstract}

@@

The four independent teams were given the task of projecting the
potential effect of increased oral cholera vaccine (OCV) efforts across Haiti.
\jwc{New sentence:}
OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.
The failure of \cite{lee20} to correctly predict the elimination of cholera has been the topic of some discussion \citep{francois20,rebaudetComment20,henrys20,leeReply20}.

Here, we demonstrate that attention to certain statistical considerations could have improved the fit to previous data, as well as leading to forecasts that match more closely the observed outcome.
\jwc{NEXT TWO SENTENCES ARE NEW: }
Because our primary focus is on the statistical tasks of fitting and diagnosing mechanistic models, we refrain from commenting on the feasibility of scientific assumptions implicitly made in \cite{lee20}.
We demonstrate that model-mispecification does not alone explain the discrepancy between model predictions and reality, as has been suggested \citep{rebaudetComment20}.
Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
We proceed by introducing Models 1--3 in Sec.~\ref{sec:models};
in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models.
In Sec.~\ref{sec:results}, we use improved model fits to project cholera incidence in Haiti under various vaccination campaigns.
We then conclude with a discussion on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:discussion}.

@@

Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ and an observable response vector $\bm{Y}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
The latent state vector $\bm{X}^{(\modelCounter)}$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R) for $\modelCounter \in \{1, 2, 3\}$, with various sub-divisions sometimes considered across the three models under consideration.
Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by $u\in \seq{1}{U}$, which corresponds here to membership of the $U=10$ administrative departments.
% Model~1 describes the dynamics aggregated at the national level.

In this study, we focus our attention on important statistical, rather than scientific, details that may improve inference on a dynamic system;
we therefore refrain from commenting on the assumptions in each model or proposing our own alternative description of cholera dynamics, and instead rely on the scientific expertise of \cite{lee20}.

@@

As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In \eqref{eq:model3:water}, $J_u(t)$ is a unit-less measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in departement $u$ during the epidemic, and $D_u$ is the average population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
We note that all deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
% Movement into each vaccination cohort is determined by previous and future vaccination campaigns.
% Here we note that only Model 3 accounted for previous vaccination efforts in the model.
% TODO: The commented lines above seemed unecessary - Jesse

@@

Many dynamic systems of scientific and public interest are highly complex by nature; therefore when developing a mathematical description of a dynamic system, it is often necessary to impose assumptions on the system in order to make inference possible.
For example, one common assumption in epidemiology is a homogeneous mixing population \citep{bansal07}.
Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
Ultimately, these decisions will result in a different approach to calibrating (or fitting) model parameters.

The models considered in this study represent a diverse selection of possible modelling decisions.
Each of the models independently describe the cholera dynamics as a Partially Observed Markov Process (POMP), since each model is structured in such a way that, given the partially-observable state of the system at time $n$, the future state of the system is independent of its past.
Because of the spatial structure of Models~2 and~3, we refer to these models as SpatPOMPs \citep{asfaw21arxiv}.
In the following subsections, we describe approaches to fitting mechanistic models and discuss potential trade-offs of the selected model complexity.

@@

One approach to modelling a dynamic system is through probabilistic models.
With a probabilistic model,
% rather than modelling the mean behavior of the system through the use of deterministic differential equations,
we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \theta)$, where $\theta$ is a parameter vector that indexes the model.
Using a probabilistic model results in several advantages, including the ability to estimate the variability of the system.
% which is of great interest to scientists and policy makers.
Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.

Once a model for the system has been proposed, the parameter vector $\theta$ needs to be estimated using the observed data.
There exist several algorithms, both frequentist and Bayesian, that can be used to obtain estimates of the parameters in stochastic dynamic models.
% including the EM algorithm, Kalman Filter (and extensions) \citep{evensen09}, and iterated filtering algorithms \citep{ionides15}.
In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
Plug-and-play algorithms include ABC and PMCMC \citep{toni09,andrieu10}, but here we focus frequentist methods to maximize model likelihoods.
To our knowledge, the only plug-and-play frequentist methods that can maximize the complete model likelihood are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.

The ability to obtain MLEs for each of the model parameters does not immediately imply that the model itself

@@

We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
Similar to Model~2, this model was originally implemented in two parts: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
As before, we continue to allow the re-estimation of some model parameters at the start of the endemic phase ($\reportRate, \sigmaProc^2$ and $\obsOverdispersion$) but require that the latent Markov process $X(t)$ carry over from one phase into the next.
The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
These results indicate that the inclusion of a trend improves the model fit to the data.
The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

@@

Parameters that must be fit in Model~3 are primarily shared between each departement, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each departement $u \in \seq{1}{10}$.
To fit this model, \citet{lee20} simplified the parameter estimation problem by fitting independent departement-level models to the data.
The shared parameters were calibrated using the cholera incidence data from Artibonite, and the departement-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective departement.
Reducing a spatially-heterogeneous model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
In particular, when considering a model for departement $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
A first order approximation of Eq.~\myeqref{eq:model3:foi} for each departement $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each departement:
\begin{equation}
  I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
\end{equation}
This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.

@@

We then fit the panel version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
% Even with the simplification of Eq.~\myeqref{eq:model3:approx}, a significant amount of computational effort is needed to obtain a proper model fit.
Therefore the reported parameters were obtained using the PanelPOMP version of Model~3, and simulations for the various vaccination campaigns (Sec.~\ref{sec:results}) were obtained using these same parameters in the SpatPOMP version of the model.
% Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.

@@

Fitting Model~3 simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
One concern is that of obtaining model projections, which was the primary goal of the original modeling exercise.
% "primary output" -> "primary goal", based on Anna's suggestion.
The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in projections.
To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package.
Simulations from the SpatPOMP version of the model, using the parameters that were fit with the PanelPOMP version of the model, are displayed in Fig.~\ref{h3spatsims}.

@@

While simulations from the SpatPOMP version of the model seem similar to the observed data, we note in Table~\ref{tab:likes} that using the MLE obtained using the approximation in Eq.~\myeqref{eq:model3:approx} results in a lower likelihood value for the SpatPOMP version of the model than the PanelPOMP version.
A natural question to ask is if the MLE obtained using the approximation in Eq.~\myeqref{eq:model3:approx} could result in a drastically different set of parameters than those that would be obtained by estimating the MLE in the fully coupled model.
Unfortunately, the answer to this question would require the ability to accurately and reliably estimate the MLE of the fully coupled model, which is beyond the scope of this article.
% this question is unsatisfactorily left unanswered, because the answer would require knowledge of the the MLE without the simplification, which is an open statistical question and beyond the scope of this article.
% While this approach may provide a reasonable estimate of the model parameters, it is unknown by how much this estimate may be improved.
This highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.

@@

\jwc{NEW PARAGRAPH/EDIT: }
The purpose of parameter calibration, irrespective of the approach taken (e.g., frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric (such as model likelihoods).
Obtaining the best set of parameters for a given model does not, however, guarantee that the model provides an approximate representation of the system in question.
Model mispecification, which may be thought of as the omission (or inclusion) of a mechanism in the model that is an important (or distracting) feature of the dynamic system, is inevitable at all levels of model complexity.
In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is often attributed; here, we assume that that some degree of model mispecification is always present, and therefore a researcher must subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".

% Researchers may use fitted models to make inference and draw conclusions about the dynamic system under investigation.
% Because of the casual nature of mechanistic models, the validity of conclusions drawn using a fitted model is dependent on the ability of the proposed model to approximate the system in question.
It is often impossible to know, however, the extent to which a model provides a faithful representation of the system.
In such cases, we rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit.
Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.

>>>>

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{\hspace{6mm} ---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{D}
\newcommand\code[1]{\texttt{#1}}
\newcommand\paramVec{\theta}

@@

\begin{abstract}
Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions for stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
% \eic{EDIT IN THE DIRECTION OF THE PREVIOUS SENTENCE.}
As a case study, we consider a cholera epidemic in Haiti.
The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission until it was eliminated in 2019.
% with the help of vaccination and other public health measures.
% \jwc{Kind of the point here is that it went away without vaccination efforts.}
We study three models developed by expert teams to advise on vaccination policies.
We assess methods used for fitting and evaluating these models, leading to recommendations for future studies.
% Methods that lead to better statistical fit include benchmarking mechanistic models against associative statistical models and using state-of-the art tools designed to fit mechanistic models.
Better statistical fit facilitates interpretation of the fitted model, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
\end{abstract}

@@

The four independent teams were given the task of estimating the
potential effect of increased oral cholera vaccine (OCV) efforts across Haiti.
\jwc{New sentence:}
OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.

The failure of \cite{lee20} to correctly predict the elimination of cholera has been the topic of some discussion \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
We demonstrate that careful attention in model fitting, diagnosing, and forecasting improves each of the proposed model's ability to quantitatively describe observed data and results in forecasts that are more consistent with the observed outcome.
\jwc{NEXT TWO SENTENCES ARE NEW: }
Because we focus on the statistical tasks of model fitting, diagnosing, and forecasting, we refrain from commenting on the feasibility of scientific assumptions implicitly made in the models under consideration.
Our results demonstrate that model-mispecification does not alone explain the discrepancy between model predictions and reality, as has been suggested \citep{rebaudetComment20}.
Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.

We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models.
In Sec.~\ref{sec:results}, we use improved model fits to project cholera incidence in Haiti under various vaccination campaigns.
We then conclude with a discussion on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:discussion}.

@@

Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ and an observable response vector $\bm{Y}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
The latent state vector $\bm{X}^{(\modelCounter)}$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model for $\modelCounter \in \{1, 2, 3\}$.
Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by $u\in \seq{1}{U}$, which corresponds to membership of the $U=10$ administrative Haitian departments (anglicized version of departements).
% Model~1 describes the dynamics aggregated at the national level.

% \jwc{Already stated above}
% In this study, we focus our attention on important statistical, rather than scientific, details that may improve inference on a dynamic system;
% we therefore refrain from commenting on the assumptions in each model or proposing our own alternative description of cholera dynamics, and instead rely on the scientific expertise of \cite{lee20}.
\jwc{NEW PARAGRAPH:}
In the following subsections, detailed descriptions of Models~1--3 are provided.
While these descriptions are a necessary inclusion for transparency and reproducibility, they do not immediately contribute to our primary goal of model fitting, diagnosing, and forecasting, with the purpose of informing policy.
A first-time reader may therefore benefit from quickly skimming through Sec.~\ref{sec:methods} and returning to the section later, as needed.

@@

As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In \eqref{eq:model3:water}, $J_u(t)$ is a unit-less measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $D_u$ is the average population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
We note that all deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
% Movement into each vaccination cohort is determined by previous and future vaccination campaigns.
% Here we note that only Model 3 accounted for previous vaccination efforts in the model.
% TODO: The commented lines above seemed unecessary - Jesse

@@

\jwc{NEW PARAGRAPH}
% Mechanistic modeling is an attempt to account for known or hypothesized features of a dynamic system.
Proposed mechanistic structures form a family of statistical models that can be indexed by a parameter vector $\paramVec$.
Because different values of $\paramVec$ result in drastically different modeling outcomes, researchers must be careful when choosing $\paramVec$.
Elements of $\paramVec$ are either chosen as constants that reflect our understanding of the system, or calibrated to observed data, with the intent of maximizing a proposed model's ability to quantitatively describe observed data according to some measure.
While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as briefly discussed in Sec.~\ref{sec:results}

The complex nature of the dynamic systems which are of scientific and public interest necessitates the imposition of various modeling assumptions in order to facilitate inference.
% Many dynamic systems of scientific and public interest are highly complex by nature; therefore when developing a mathematical description of a dynamic system, it is often necessary to impose assumptions on the system in order to make inference possible.
For example, one common assumption in epidemiology is a homogeneous mixing population \citep{bansal07}.
Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
Ultimately, these decisions may result in a different approach to calibrating model parameters.

\jwc{MAJOR CHANGE TO THIS PARAGRAPH:}
Each model considered in this study describes cholera dynamics as a Partially Observed Markov Process (POMP), as each model contains unobservable states that evolve sequentially with time.
Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
In the following subsections we describe our approach to fitting these mechanistic models.

@@

One approach to modeling a dynamic system is through probabilistic models.
With a probabilistic model,
% rather than modeling the mean behavior of the system through the use of deterministic differential equations,
we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
Using a probabilistic model results in several advantages, including the ability to account for variability present in the system.
% which is of great interest to scientists and policy makers.
Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.

Once a model for the system has been proposed, the parameter vector $\paramVec$ needs to be estimated using the observed data.
There exist several algorithms, both frequentist and Bayesian, that can be used to obtain estimates of the parameters in stochastic dynamic models.
% including the EM algorithm, Kalman Filter (and extensions) \citep{evensen09}, and iterated filtering algorithms \citep{ionides15}.
In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use frequentist methods to maximize model likelihoods.
To our knowledge, the only plug-and-play frequentist methods that can maximize the complete model likelihood are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.

@@

We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
Similar to Model~2, this model was originally implemented in two parts: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
As before, we continue to allow the re-estimation of some model parameters at the start of the endemic phase ($\reportRate, \sigmaProc^2$ and $\obsOverdispersion$) but require that the latent Markov process $X(t)$ carry over from one phase into the next.
The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
These results are suggestive that the inclusion of a trend in transmission rate improves the quantitative ability of Model~1 to describe the observed data.
The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

@@

Parameters that must be fit in Model~3 are primarily shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
To fit this model, \citet{lee20} simplified the parameter estimation problem by fitting independent department-level models to the data.
The shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
Reducing a spatially-heterogeneous model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
A first order approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
\begin{equation}
  I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
\end{equation}
This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.

@@

We then fit the panel version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
% Even with the simplification of Eq.~\myeqref{eq:model3:approx}, a significant amount of computational effort is needed to obtain a proper model fit.
The reported parameters were obtained using the PanelPOMP version of Model~3, and simulations for the various vaccination campaigns (Sec.~\ref{sec:results}) were obtained using these same parameters in the SpatPOMP version of the model.
% Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.

@@

Fitting Model~3 as a PanelPOMP simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
% "primary output" -> "primary goal", based on Anna's suggestion.
The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package.
Simulations from the SpatPOMP version of the model, using the parameters that were fit with the PanelPOMP version of the model, are displayed in Fig.~\ref{h3spatsims}.

@@

While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
A natural question to ask is if the MLE obtained for the PanelPOMP version of Model~3---obtained using the approximation in Eq.~\myeqref{eq:model3:approx}---could result in a drastically different set of parameters than those that would be obtained by estimating the MLE in the fully coupled model.
Unfortunately, the answer to this question would require the ability to accurately and reliably estimate the MLE of the fully coupled model, which is beyond the scope of this article.
% this question is unsatisfactorily left unanswered, because the answer would require knowledge of the the MLE without the simplification, which is an open statistical question and beyond the scope of this article.
% While this approach may provide a reasonable estimate of the model parameters, it is unknown by how much this estimate may be improved.
This highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.

@@

\jwc{NEW PARAGRAPH/EDIT: }
The purpose of parameter calibration, irrespective of the approach taken, frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric.
Obtaining the best set of parameters for a given model does not, however, guarantee that the model provides an approximate representation of the system in question.
Model mispecification, which may be thought of as the omission (or inclusion) of a mechanism in the model that is an important (or distracting) feature of the dynamic system, is inevitable at all levels of model complexity.
In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is often attributed: we assume that some degree of model mispecification is always present, and it is necessary to subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".
It is often impossible to know, however, the extent to which a model provides a faithful representation of the system in question.
In this section, we discuss some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective by relying on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit.
Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
